---
title: "Architecture: how the backend registry works and how a new engine joins"
vignette: >
  %\VignetteIndexEntry{Architecture: how the backend registry works and how a new engine joins}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
  %\usepackage[utf8]{inputenc}
output:
  rmarkdown::html_vignette
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  eval     = FALSE,
  purl     = FALSE
)
library(flexyBayes)
```

This vignette is for readers who want to understand how flexyBayes
decides what each backend can do, and what it would take to add a new
backend. It describes the package's internal architecture; the surfaces
shown here are intentionally not part of the everyday user API, so the
code is illustrative rather than something a typical analysis calls.

## Backend as a first-class axis

The package is organised around a small number of closed-vocabulary
registries -- for representations, approximation schemes,
backend-independence pairs, refusal reason-codes, and backends. Each is
a locked table established when the package loads. A registry is the
single home for one kind of fact, so the rest of the package reads from
it rather than hard-coding the same knowledge in several places.

The backend registry is the home for backend facts: the legitimate
backend names, what grammars each can ingest, which inference paradigm
each uses, which R package makes each usable, whether each is in the
`auto` candidate set, and which models each can represent. Dispatch
*consumes* these facts; it does not restate them.

```{r registry-facts}
# The active backends whose required package is installed.
flexyBayes:::.available_backend_names()

# The candidate set that `auto` considers by default.
flexyBayes:::.auto_default_backend_names()
```

The design rule is *extend by registration, not by API growth*: new
capability enters by adding a row to a registry, not by widening the
public verb surface. The everyday surface -- `fb()` / `flexybayes()`
plus the three engine pins -- does not change when a backend is added.

## What an entry records

Each backend entry carries a fixed set of fields. The two that govern
behaviour most directly are the inference `paradigm` and the
`capability_predicate`.

The `paradigm` is the honest label for how a backend draws inferences --
for example, full Hamiltonian Monte Carlo versus a Laplace
approximation. It is what lets `triangulate()` and the
backend-independence registry report whether two fits are genuinely
independent: two engines that share a paradigm agree for weaker reasons
than two engines that do not. The brms backend keeps `brms` as its
front-end name while recording its sampler paradigm separately, so a
greta-and-brms comparison is graded as same-paradigm rather than being
mistaken for independent evidence.

The `capability_predicate` is a function of the model that returns
either "this engine can represent it" or a single reason-code naming why
it cannot. This is what makes gating *systematic* rather than a
collection of special cases. For example, the brms predicate refuses an
ASReml structured-covariance term, because Stan has no lossless
translation for it:

```{r capability}
d <- data.frame(
  y    = rnorm(40),
  env  = factor(rep(letters[1:4], each = 10L)),
  geno = factor(rep(seq_len(10L), times = 4L))
)
ir <- fb_from_asreml(y ~ env, random = ~ geno, data = d)

# greta can fit a plain random-intercept model; so can brms.
flexyBayes:::.backend_can_fit("greta", ir)
flexyBayes:::.backend_can_fit("brms",  ir)
```

A predicate delegates to the existing authority for its decision rather
than duplicating it -- the INLA predicate, for instance, asks the
latent-Gaussian gate -- so a model can never be judged feasible by one
part of the package and infeasible by another.

## Transparency by construction

"All the backends the model supports" is an honest claim because every
unsupported (model, engine) pair yields a registered, structured
refusal. The refusal registry holds the closed vocabulary of
reason-codes, and every user-facing refusal is drawn from it, so a
refusal can always be matched on its code rather than parsed from free
text.

```{r refusals}
# The registered refusal vocabulary (machine-readable reason-codes).
fb_refusals()
```

On the `auto` path, the engines that were considered and rejected are
recorded on the fit and surfaced by `backend_decision()`, so a user can
always see why a particular backend was chosen.

## What it takes to add a backend

Adding a backend that uses an inference paradigm already present comes
down to two things: registering the backend (its name, grammars,
paradigm, required package, `auto` membership, and capability predicate)
and providing an emit function that lowers the intermediate
representation to that engine's code. Dispatch resolves the emit
function from the registry by name, so no call site needs to hard-code
the new engine.

A backend that introduces a *new* inference paradigm needs one more
thing: an orchestration step, because the per-paradigm routing order and
fallback policy -- which engine is the fast path, which is the universal
fallback, which is opt-in only because of compile latency -- is genuine
policy rather than mechanical iteration, and the package keeps it as
explicit code rather than deriving it from the registry.

The roster has room for growth by design. A backend slot can be
provisioned in a dormant state -- registered but not yet active, so it
refuses cleanly at dispatch until activated -- which is how a new
front-end can be reserved before its emit path is ready.

## Summary

- Backend facts live in one locked registry; dispatch reads from it.
- An entry records the inference `paradigm` (so `triangulate()` grades
  independence honestly) and a `capability_predicate` (so gating is
  systematic and delegates to existing authorities).
- Every unsupported pair yields a registered, structured refusal;
  rejected routes are recorded on the fit.
- A same-paradigm engine is added by registration plus an emit hookup; a
  new paradigm additionally needs an orchestration step.
