---
title: "Tutorial 13: Per-term INLA memory feasibility"
date: "2026-05-28"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Tutorial 13: Per-term INLA memory feasibility}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>"
)
```

# 1. Why a per-term memory estimator?

`lgm_gate()` rule 11 (`memory_feasibility_inla`) originally used a
single multiplier on the preflight's indexed estimate. That multiplier
was a useful safety net but coarse: a model whose memory profile is
dominated by one structured-covariance term could refuse for the wrong
reason, and the diagnostic message named neither the binding term nor
the representation class driving the projection.

The gate now uses a per-term decomposition instead. Each random-effect
term contributes bytes according to its representation class:

| Representation | Bytes per term | Notes |
|---|---|---|
| `indexed_random_intercept` | `8 K + 4 N` | K-vector of latent u's + integer index column. |
| `indexed_structured_known` (precision) | `nnz(Q) * 16` | Sparse-matrix bytes-per-non-zero from Lindgren et al. (2011); falls back to `K * 16` when `known_matrices` is absent. |
| `block_diagonal` | `sum_k n_k^2 * 8` | Block-diagonal V across K independent blocks; falls back to `K^2 * 8` worst case when `known_matrices` is absent. |
| `dense_baseline` | (already in design_memory_bytes) | Dense N x p basis blocks. |

The fixed-effect contribution is the design matrix's `N x p * 8`
bytes. The grand total is multiplied by an overhead factor (default
2.0; configurable via the
`flexyBayes.preflight_inla_overhead_factor` option) that absorbs
INLA's internal stage-2 allocations.

# 2. How to read the breakdown

`fb_plan()` and `flexybayes(plan = TRUE)` surface the breakdown
under the headline memory line of the plan's print form. The
underlying data sits on `plan$preflight$memory_estimate`:

```
List of 4
 $ total          : num <total bytes after overhead>
 $ breakdown      :'data.frame': <K + 1> obs. of  4 variables:
  ..$ term_label    : chr  <one row per random term + "(fixed effects)">
  ..$ representation: chr  <representation class>
  ..$ bytes         : num  <bytes>
  ..$ share         : num  <share of raw total>
 $ overhead_factor: num 2
 $ raw_total      : num <bytes before overhead>
 - attr(*, "class") = c("fb_memory_estimate", "list")
```

The shape is stable: the field names follow a closed vocabulary.
`as.numeric(plan$preflight$memory_estimate)` returns the total in
bytes, so existing scalar-numeric consumers continue to work
unchanged.

# 3. How to negotiate a memory refusal

When the per-term estimator's total exceeds the active ceiling,
`lgm_gate()` refuses with `memory_feasibility_inla_per_term` and
names the largest contributor on the condition's diagnostic line.
Four common patterns resolve the refusal:

1. **Route via greta.** `backend = "greta"` keeps the indexed
   representation throughout MCMC; the dense materialisation INLA's
   stage-2 needs is bypassed entirely.
2. **Reduce structured-cov cardinality.** Aggregate levels of the
   grouping factor when the science permits (e.g., site x season
   instead of site x date).
3. **Switch dense to sparse precision.** If the structured covariance
   has known sparse precision, pass `precision = solve(V)` instead
   of `V` so the term routes through INLA's sparse generic0 path.
4. **Raise the ceiling.** Pass `memory_ceiling_gb` (or set
   `flexyBayes.preflight_ram_fraction`) when the machine genuinely
   has the headroom and the per-term breakdown reads honestly.

# 4. What the per-term estimator does not capture

The per-term estimator covers the four exact representations
flexyBayes supports today (dense, chol, precision/pedigree-sparse,
blocks) plus the indexed random-effect and dense-basis fixed-effect
shapes. Two surfaces are explicitly out of scope:

- **SPDE meshes.** INLA's `model = "spde2"` mesh allocations are
  workflow-specific and not modelled here; a future release
  ships a tensor-smooth estimator that covers the relevant shapes.
- **Approximate schemes** (low-rank covariance carriers, variational
  stage-2). These refuse at the routing layer until validated
  through `validate_approximation()`.

# 5. References

Lindgren, F., Rue, H., Lindstrom, J. (2011). "An explicit link
between Gaussian fields and Gaussian Markov random fields: the
stochastic partial differential equation approach." *Journal of
the Royal Statistical Society, Series B*, 73(4), 423-498.

Rue, H., Martino, S., Chopin, N. (2009). "Approximate Bayesian
inference for latent Gaussian models by using integrated nested
Laplace approximations." *Journal of the Royal Statistical
Society, Series B*, 71(2), 319-392.

```{r sessioninfo, echo = FALSE}
sessionInfo()
```
